An Observational Study to COVID-19 and 2020 US Presidential Election: Counties with High Deaths per Case Rate Showed Diminished Support For Trump

### 2023-12-15 outputs/paper/paper.qmd
Update the final paper and references

---
title: "An Observational Study to COVID-19 and 2020 US Presidential Election: Counties with High Deaths per Case Rate Showed Diminished Support For Trump"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss.git"
date: "today"
date-format: "long"
abstract: "This study investigate the causal inference between COVID-19 and Donald Trump's loss during the 2020 US Federal Election using the data from MIT Election Data Science Lab and Johns Hopkins University CSSE. The main methodology used in this paper is Propensity Score Matching with with an exploratory analyzing the optimal treatment initially. The key finding suggest that the counties with a death per case rate exceeding the 0.4 quantile threshold shows a reduced voting preference for Trump. Additionally, this paper conducts a counterfactual analysis based on the treatment effect, indicating that Trump might have be secured to re-elect if the disparities in death per case rates were addressed. Future research should aim to refine the propensity score calculations by incorporating additional variables, such as the winning party in each county during the 2016 election."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

At the beginning of 2020, US reported the first COVID-19 cases. This number grows to ten thousand within two months on March 19 and increase dramatically to 100K on March 27 [citation]. At the same year, the U.S. Federal Presidential Election started off and Trump lost to Biden. As of Election Day, there was above 9 million cases and about 0.2 million deaths in US. When people talk about Trump's lost, people always connect it to the pandemic, saying that he "mishandled his greatest test" [citation] and 2016 Election was a historical accident [citation]. However, there is another voice that COVID is not the only or the decisive factor attributing to his lost, his limits as a politican and wasted his advantage on the economy are also important explaining why he lost. This paper will investigate whether the is a causal effect between COVID and Trump's lost. If there is, whether it is strong enough to beath Trump at 2020.

There are existing researches on this. Leaonardo et al. [citation] suggested that Trump could win if there was no COVID but it will be too naive to simply connect his lost to the COVID-19 infection rate. Their research highlighted that the COVID-19 had the most significant negative impacts on those urban areas where there was no stay-at-home orders, especially for those "swing" states. They also argue that the race diversity and education attainment are also important when analysing voting patterns for Trump. In addition to that, Marcus et al. [citation] suggested that the deaths per case is a more crucial indicator than infection rate when analysing the voting for Trump. However, they suggest the voting itself may not capture the impact of COVID as we do not know when did the voters make their decision, especially for early vote. Compared to these researches, Harold [citation] conduct read survey before and after the election to illustrate the impact of COVID pattern, they found that even though COVID has impacts on voting, it is not the dominant factor. Without COVID, Trump was also less likely to re-elect due to US highly polarized political landscape. Similarly, Shang et al. [citation] also conduct a survey and all the participants show natural or negative altitude. They suggest some social activity like "Black lives matter" may have impacts on voting. Furthermore, the

While all the above researches had investigated the correlation between COVID-19 and votings for Trump. Except conducting real surveys, the methods they used were to conduct a model and find the relevant variables in predicting the (change) vote for Trump. Then conduct counterfactual analysis to predict what will be the voting for Trump if there was no COVID or a certain amount reduce in COVID deaths. Instead using a similar way like those studies, this paper will introduce the method of Propensity Score Matching to find the treatment effect, to see whether those more impacted counties will have different voting patterns for Trump. However, since all counties experienced COVID-19, this paper will firstly find the optimal treatment groups. After that, I will also conduct a counterfactual analysis to see if Trump could re-elect.

There will be five parts in this paper. I will introduce the data used in this paper and present data suammries and visualization in [Data] part. After that, I will introduce the methods in this paper and the corresponding results in  [Methods] and [Results] part, respectively. Then the results will be interpreted and discussed in the [Discussion] part and conclude with limitations and drawbacks in [Conclusion] part..

# Data

@@

The data used in this paper comprised five data sets from three different sources corresponding to different topics. The primary source is MIT Election Data Science Club which build open online data collections of the US Federal or Senate Election results, spanning from nation to county levels. The data extracted is called "County Presidential Election Returns 2000-2020" with about 70,000 rows containing the voting patterns for each candidate and party by county since 2000. Besides that, the data also indicate the types of voting, such as "EARLY VOTE" and "ELECTION DAY," for the same party in the a county. To analyze the voting patterns for Donald J. Trump, I only select the data for 2016 and 2020 US Federal Election for all counties and parties. Thus the filterd data set contains the number of votes through various ways for each county and party.

In addition to that, the study also uses the data regarding COVID-19 from the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University. CSSE collects and reports the local, national, and global multidimensional data, including medicine, health care, disaster response, etc. During the pandemic, they collected the U.S. and international COVID cases and deaths and reported them on their GitHub, summarized by daily reports from April 12, 2020, to March 9, 2023. To access the impact of COVID on 2020 Election, this paper used the daily report on November 3, 2020 which is the election day of 2020 US Federal Election. The resulting data include the aggregates number of cases, deaths and recovers, as well as the incidence rate and case fatility ratio for each county by the election day.

Lastly, this paper also employ the socio-economic data from American Community Survey (ACS). ACS is a online open source database conducted by the U.S. Census Bureau, containing the various soci-economic factors in either geographic levels. The data sets extracted from ACS is 2020 five-year estimates of DP02, DP03 and DP05 which covers the social, economic and demographic characteristics in county level. Since there are thousands of variables, I will only use the variables which are found from my previous paper that are significant in predicting the COVID-19 mortality rate such as the mean household income, high-education attainment etc. The descriptions of all variables can be found on Table 1.

After having five data sets from three sources, they are merged into one big data by counties. Using the merged data, there are some new variables calculated by the exiting variables. In terms of election, I calculate the percentage of vote for each party (candidate) in both 2016 and 2020 Election, and calculate the corresponding change of percentage votes. I also create new variables indicating the winning party in both elections for each county. Additionally, to accurately compare the COVID cases and deaths across all counties, I transform the number of cases and deaths to infection and mortality rate by 10,000 citizens in each county. The final data now consists of 3107 rows with 36 columns. All the important variables are described in Table 1.

```{r}
#| tbl-cap: Descriptions of all important variables in the analyze data
#| label: tbl-descriptions
data.frame(var = c("Income Percentile", "High-Education Attainment",
                   "Private Insurance", "No Insurance", "White Population",
                   "Black Population", "Males Population", "Infection Rate",
                   "Mortality Rate", "Death per Case", "Votes for Democrat (2016)",
                   "Votes for Democrat (2020)", "Votes for Republican (2016)",
                   "Votes for Republican (2020)", "Change of Vote for Republican"),
           code = c("income_pctile", "prop_high_education", "private_insurance",
           "no_insurance", "white_pct", "black_pct", "males", "infrate",
           "mortrate", "dpc", "pct_vote_demo16", "pct_vote_demo", 
           "pct_vote_rep16", "pct_vote_rep", "change_vote_rep"),
           descriptions = c("The income percentile of each county",
                            "The proportion of residents in a county having a at least bachelor degree",
                            "The proportion of local residences having private insurance",
                            "The proportion of local residences without any health insurance",
                            "The proportion of White population",
                            "The proportion of Black population",
                            "The proportion of Males population",
                            "The COVID infection rate, calculated by number of case per 10,000 residences",
                            "The COVID mortality rate, calculated by number of death per 10,000 residences",
                            "The COVID death per 10,000 confirmed cases",
                            "The percentage of votes for the Democrat in 2016",
                            "The percentage of votes for the Democrat in 2020",
                            "The percentage of votes for the Republican in 2016",
                            "The percentage of votes for the Republican in 2020",
                            "The change of vote for the Republican from 2016 to 2020")) %>% 
  rename(`Variables` = var,
         `Coded name` = code,
         `Descriptions` = descriptions) %>% 
  kable(booktabs = TRUE)
```

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(n = n(),
            case = mean(cases),
            inf = mean(infrate),
            deaths = mean(deaths),
            mort = mean(mortrate),
            dpc = mean(dpc),
            income = mean(mean_household_income)) %>% 
  rename(`Winning Party` = winning_party,
         `Count` = n,
         `Cases` = case,
         `Infection Rate` = inf,
         `Deaths` = deaths,
         `Mortality Rate` = mort,
         `DPC` = dpc,
         `Income` = income) %>% 
  kable(booktabs = TRUE)
```

@tbl-covid summarize the COVID-19 impacts and income levels between the counties voting for Democrat and Republican. Even though the Republican won about five sixths counties, there is a stark contrast where counties supporting Biden had nearly ten times the average number of COVID-19 cases and deaths compared to those supporting Trump. Despite a similar infection and mortality rates between the two groups, there is still a notable higher death-to-case rate in counties voting for the Republican. Furthermore, these counties also has a lower mean household income than the "Democrat-win" counties. These patterns suggest a potential correlation between the extent of COVID-19 impact and voting behavior in the 2020, where areas more severely affected by COVID-19 were less likely to vote for the Trump, vice-versa. This aligns with mainstream media and intuitive expectations.

@@

p1 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = pct_vote_rep), color = "black") +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "grey") +
  labs(title = "The relative share of votes between Republican and Democrat party",
       fill = 'Share of Republican') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.9)

@@

We can validate our guess from @fig-map which compares the relative ratio of votes for the two parties and the death to cases rate in maps. From the maps, it is clearly more counties had preference for the Republican but the counties with higher death per case rate have a smaller ratio of vote for the Republican and these counties are usually richer ones. The most significant example can be California and New York. In contrast, the states with less impacted by COVID, such as Utah, vote more for Republican.

@@

Given the insights from @tbl-covid and @fig-map, @fig-dpc-income further examine the impact of COVID and income levels on the voting behaviors for the two parties. @fig-dpc-income shows the correlation between the votes of the two parties to the death per case rate and income levels. The two graphs on the left indicates the share of votes whereas the right two graphs indicate the number of votes. The two parties are represented by different colors. In terms of share of votes, in general, there is a positive relationship between the DPC and votes for the Democrat but a negative one for the Republican. However, the counties with medium value of DPC is less sensitive to the voting for the two parties, compared to the "tail" counties which their support change significantly as DPC increases or decreases. Meanwhile, there is a quadratic patterns between the vote for the two parties and income, it seems that the Republican is in favor of low income counties, but as income increase, more counties vote for Bide not Trump.

In addition, even though the average share of votes for Republican is always higher than 0.5, it does not necessarily mean that the Democrat lost in all counties. The reason behind this is, as mentioned in @tbl-covid, the counties which the Republican won is about five times as higher as the Democrat. Since Republican has more counties, it definitely has a higher average share of votes than Democrat. This pattern also indicate that Democrat usually won the densely populated, urban areas but Republican won in more sparsely populated, rural areas \[citation\]. The high population density areas has more votes and explains why Biden beat Trump. We can verify this from the number of votes for the two parties, where the number of votes for Biden exceeds Trump as DPC and income level increases. This pattern is especially significant for the top 50% richest counties and the counties with DPC larger than 75 quantile.

```{r}
data_population <- data %>% 
  mutate(pop_pctile = ntile(total_population, 100)) %>% 
  group_by(pop_pctile) %>% 
  summarise(county_demo = sum(winning_party == "Democrat")/n(),
            county_rep = sum(winning_party == "Republican")/n())

data_population %>% 
  ggplot(aes(x = pop_pctile)) +
  geom_point(aes(y = county_rep, colour = "Republican"), alpha = 0.25) +  
  geom_smooth(aes(y = county_rep, colour = "Republican"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  geom_point(aes(y = county_demo, colour = "Democrat"), alpha = 0.25) + 
  geom_smooth(aes(y = county_demo, colour = "Democrat"), 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Party",  
                      labels = c("Democrat", "Republican")) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  labs(x = "Population Percentiles", 
       y = "Share of winning counties")
```

```{r}
boundaries <- c(0, 5000, 9000, 14000, 19000, 26000, 37000, 54000, 94000, 215000, Inf)

# Classify the counties into percentiles
data_population <- data %>% 
  mutate(pop_pctile = cut(total_population, breaks = boundaries, include.lowest = TRUE, labels = FALSE))

# data_population %>% 
#   filter(winning_party == "Republican") %>% 
#   ggplot(aes(x = pop_pctile)) +
#   geom_bar(aes(y = n()))
# 
# data_population %>% 
#   group_by(pop_pctile, winning_party) %>% 
#   summarise(n = n())
# 
# county_sum <- data_population %>%
#   filter(winning_party == "Republican") %>%
#   group_by(pop_pctile) %>%
#   summarize(count = n())

# Plot the data
# ggplot(county_sum, aes(x = pop_pctile, y = count)) +
#   geom_bar(stat = "identity")
```

From four graphs in @fig-dpc-income, we can conclude that, in general, the higher the DPC rate, the lower the support for Trump but higher for Biden. Besides, the Democrat is in favor of richer voters, where as Republican is in favor of low income counties. It is direct to ask whether these two variables have interaction effect on the voting patterns? However, it will be difficult to visualize the interaction effect between two continuous factors. Instead, I will manually set a new categorical variable `high_income` to denote whether the county has a high income with certain cutoff. Then the difference of correlations of the two groups as death per case rate changes. I will set different cutoffs to fully test the interaction effect.

```{r}
#| echo: false
#| message: false
#| label: fig-boxplot
#| fig-cap: Summary of income levels for counties voting for Democratic and Republican

data  %>%
  ggplot(mapping = aes(x = winning_party, y = mean_household_income)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.04, width = 0.3, height = 0) +
  theme_minimal() +
  labs(x = "Winning Party",
       y = "Mean Household Income")
```

```{r, fig.width=10, fig.height=4.5}
#| echo: false
#| message: false
#| label: fig-income_state
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: Variation in COVID-19 Mortality Rates Across States, Categorized by Income Levels and Dominant Political Preferences

# Find the wining data for each state
data_win_party <- data %>%
  group_by(state) %>%
  summarise(dem_votes = sum(vote_demo),
            rep_votes = sum(vote_rep), 
            .groups = 'drop') %>% 
  mutate(winning_party_state = ifelse(dem_votes > rep_votes, 'Democratic', 'Republican')) %>% 
  dplyr::select(state, winning_party_state)
# Calculate the average income
average_income_by_state <- data %>%
  group_by(state) %>%
  summarise(average_income = mean(mean_household_income), .groups = 'drop')
# Join the winning party and average income back to the original data
data_win_party_income <- data %>%
  left_join(data_win_party, by = "state") %>%
  left_join(average_income_by_state, by = "state")
# Arrange states by average income
ordered_states <- data_win_party_income %>%
  arrange(average_income) %>%
  .$state %>%
  unique()
# Create a factor with levels based on income-ordered states for plotting
data_win_party_income$state <- factor(data_win_party_income$state, levels = ordered_states)

average_dpc <- mean(data_win_party_income$mortrate)

ggplot(data_win_party_income, aes(x = state, y = mortrate, fill = winning_party_state)) +
  geom_boxplot() +
  geom_hline(yintercept = average_dpc, linetype = "dashed", color = "black") +
  annotate("text", x = Inf, y = average_dpc, label = " National Average", hjust = 1, vjust = -0.5, color = "black", size = 3) +
  scale_fill_manual(values = c('Democratic' = 'blue', 'Republican' = 'red')) +
  labs(y = 'Death Rate', x = 'State (Ascending Income)', 
       fill = "Winning Party") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

@fig-boxplot compares the distribution of county income levels for the two parties. In general, counties voting for the Democrats have higher income levels than the Republicans. In addition, we can observe that the counties are intensively concentrated around income levels \$50k to \$75k, compared to the Democrats, where the counties are approximately uniformly distributed at each income level. Furthermore, barely any county has at least 150k voting for the Republican party. Both @tbl-covid and @fig-boxplot indicates that the Democrat is in favour of wealthy counties but poorer counties for Republican. Consequently, the more affluent counties (states) usually have more electoral votes than the poorer ones; this may provide insights into why Trump lost.

# Methods

The objective in this paper is to conduct the causal inference between COVID-19 and Trump's loss during 2020, the main methods implemented in this study will be the Propensity Score Matching (PSM). After find the causal inference, this paper will also conduct a counterfactual analysis to see whether Trump can re-elect if there was no COVID.

@@

The treatment refers to the intervention or exposure that is being studied to understand its potential impact on the outcome. In other words, we implement certain interventions to a group of people but not for the rest. The group of people receiving the treatment is called the treatment group and control group for the rest. For instance, Jochen \[citation\] examined whether a new medical treatment called clampless off-pump coronary artery bypass (clampless OPCAB) will reduce the in-hospital mortality rate, compared to the traditional way. In their studies, the treatment is the new medical surgery and patients who took this surgery is the treatment group. They compared the mortality rates between the two groups and concluded that the new treatment can significantly lower the stroke and mortality rate.

Normally, the treatment should be established before conduct the analysis, this aims to reduce the risk of p-hacking. However, the objective of this paper is to find whether Trump lost the last Election due to COVID-19, it may not be possible to state the treatment in advance as all the counties in US had experienced COVI-19. Therefore this paper will firstly perform exploratory analysis to firstly find the treatment group which has the most significant difference in voting for Trump. This paper will set a cutoff defining whether a county has a high or low death per case rate. In addition, from @fig-dpc-income, low and high income counties also seem to have different voting behaviors. Therefore the treatment in this paper will incorporate whether a county has a high DPC and a high income levels with certain cutoffs.

>>>>

---
title: "An Observational Study to COVID-19 and 2020 US Presidential Election: Counties with High Deaths per Case Rate Showed Diminished Support For Trump"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss.git"
date: "today"
date-format: "long"
abstract: "This study investigates the causal inference between COVID-19 and Donald Trump's loss during the 2020 US Federal Election using the data from MIT Election Data Science Lab and Johns Hopkins University CSSE. The main methodology used in this paper is propensity score matching with an exploratory analysis of the optimal treatment. The key finding suggests that the counties with a death per case rate exceeding the 0.4 quantile threshold show a reduced voting preference for Trump. This paper also conducts a counterfactual analysis based on the treatment effect, indicating that Trump might have been secured to re-elect if the disparities in death per case rates were addressed. Future research should refine the propensity score calculations by incorporating additional variables, such as the winning party in each county during the 2016 election."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

In early 2020, the United States encountered its first cases of COVID-19. The situation escalated rapidly, with reported cases reaching 10,000 by March 19 and soaring to 100,000 just eight days later [@WikiCOVIDUSA2023]. Amidst this unfolding crisis, the U.S. Presidential Election was underway, ending with Trump's defeat to Biden. By Election Day on November 3, there were more than 9 million reported cases and about 200,000 deaths in the U.S. When talking about Trump's loss, the public refers to his mishandling of "his greatest test" [@Greenblatt2021] and believes that his win in the 2016 Election was a historical accident [@Bryant2020]. Conversely, some arguments suggest that COVID-19 was not the sole or decisive factor in Trump's defeat; his limits as a political strategy and wasting his advantage on the economy also contributed to his loss. This paper will investigate the causal effect between COVID and Trump's loss, and if so, to what extent it influenced the 2020 election outcome.

Existing research on this subject provides varied perspectives. Baccini et al. [@Baccini2021CovidElection] suggested that Trump could win without COVID, but simply using the COVID-19 infection rate as the only factor contributing to his loss would be too naive. Socio-economic factors like high education attainment and race diversity are also unignorable. Besides, their research emphasized the significant impact of COVID-19 on those urban areas where there were no stay-at-home orders, particularly for the "swing" states. In addition, Noland & Zhang [@NolandZhang2021] suggested that the deaths per case is a more crucial metric than the infection rate when analyzing the impact of COVID-19 on voting for Trump. However, they highlight the challenge in assessing COVID-19's impact on voting, as we may not know when the voters made their decisions. In contrast, Clarke et al. [@Clarke2021CovidTrump] conducted actual surveys before and after the election, suggesting that COVID-19 impacts voting but not the dominant one. The U.S.'s highly polarized political landscape might be more significant. Similarly, Hart [@Hart2021CovidTrump] also conducted a survey to ask about people's attitudes toward Trump. All the participants show natural or negative possession. They also find that some social movements like "Murder of George Floyd" may impact voting.

While all the above research investigated the correlation between COVID-19 and voting for Trump, previous studies predominantly employed models to identify variables influencing Trump's vote share and perform counterfactual analyses to estimate his voting performance in the absence of COVID-19 or reduction in deaths. Instead of using that method, this study employs propensity score matching to find the treatment effect and analyze whether counties more severely affected by COVID-19 exhibit distinct voting patterns. However, given all counties have experienced COVID-19, this paper will first find the optimal treatment group(s). Finally, I will conduct a counterfactual analysis to see if Trump could be re-elected with the treatment effect eliminated.

This paper will have five parts. I will introduce the data used in this paper and present data summaries and visualization in the @sec-data part. After that, I will explain the methods in this paper and the corresponding results in the @sec-methods and @sec-results parts, respectively. All the results will be interpreted and discussed in the @sec-discussion part, and I will conclude with limitations and drawbacks in the @sec-conclusion part.

# Data{#sec-data}

The data in this paper is either downloaded directly or accessed via API. `R` [@R] will be the computer language used in the paper. Moreover, this paper will also use different R packages. The data is cleaned using the package `dplyr` [@dplyr], `stringr` [@stringr], `tidyverse` [@tidyverse], `janiotr` [@janitor], `tools` [@tools] and `tidyr` [@tidyr]. In addition to that, `ggplot2` [@ggplot2], `RColorBrewer` [@RColorBrewer], `maps` [@maps], `mapdata` [@mapdata], `gridExtra` [@gridExtra] and `cowplot` [@cowplot] will be used later to make tables and plot graphs. Besides, to perform Propensity Score Matching, this paper will also employ `Matching` [@Matching].

@@

This paper comprised five data sets from three different sources, each corresponding to different topics. The primary data is from the MIT Election Data Science Club [@MITElectionLab], which builds open online data collections of the U.S. Federal or Senate Election results from national to county levels. The data extracted is called "County Presidential Election Returns 2000-2020," [@MITElectionData2022] with about 70,000 rows containing the voting patterns for each candidate and party by county since 2000. Besides that, the data also indicate the types of voting, such as "EARLY VOTE" and "ELECTION DAY." To analyze the voting patterns for Donald J. Trump, I only filter the 2020 U.S. Federal Election data for all counties and parties.

Additionally, the data on COVID-19 is taken from the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University [@jhu]. CSSE collects and reports local, national, and global multidimensional data, including medicine, health care, disaster response, etc. During the pandemic, they collected the U.S. and international COVID cases and deaths, reporting them by daily summary on their GitHub [@CSSECOVID19Data]. To examine the impact of COVID-19 on the 2020 Election as accurately as possible, this paper used the daily report on November 3, 2020, which is the election day of the 2020 U.S. Federal Election. The resulting data include the aggregate number of cases, deaths and recovers before Election Day and the incidence rate and case-fatality ratio for each county.

This paper also incorporates the socioeconomic data from the American Community Survey (ACS) [@acs], which is an online open-source database conducted by the U.S. Census Bureau, containing the various socioeconomic factors at different geographic levels. The data sets extracted from ACS are 2020 five-year estimates of DP02, DP03 and DP05, covering the social, economic and demographic characteristics at the county level. Since there are thousands of variables, to ensure simplicity, I will only use the variables that are found to be significant in predicting the COVID-19 mortality rate from my previous paper [@CaoCOVID19Repo]. The descriptions of all variables can be found on @tbl-descriptions.

After having five data sets, they are merged into one big data by county. Using the merged data, I calculate the percentage of votes for each party (candidate) in the 2020 election. In addition, I also create new dummy variables indicating the winning party in both elections for each county. Moreover, to accurately compare the COVID cases and deaths across all counties, I transform the number of cases and deaths to infection and mortality rate by 10,000 citizens in each county. Using the conclusion from Noland & Zhang's work [@NolandZhang2021], I also calculate each county's deaths per case rate. The final data consists of 3107 rows with 36 columns. All the essential variables are described in @tbl-descriptions.

```{r}
#| tbl-cap: Descriptions of all important variables in the analyze data
#| label: tbl-descriptions
data.frame(var = c("income_pctile", "prop_high_education", "private_insurance",
           "no_insurance", "white_pct", "black_pct", "males", "infrate",
           "mortrate", "dpc", "pct_vote_demo", "pct_vote_rep"),
           source = c("ACS", "ACS", "ACS", "ACS", "ACS", "ACS", "ACS", "JHU", 
                      "JHU", "JHU", "MIT", "MIT"),
           descriptions = c("The mean household income percentile of each county",
                            "The proportion of local residents having a at least bachelor degree",
                            "The proportion of local residences having private insurance",
                            "The proportion of local residences without any health insurance",
                            "The proportion of White population",
                            "The proportion of Black population",
                            "The proportion of Males",
                            "The COVID infection rate, calculated by cases per 10,000 residences",
                            "The COVID mortality rate, calculated by deaths per 10,000 residences",
                            "The COVID death per 10,000 confirmed cases",
                            "The percentage of votes for the Democrat in 2020",
                            "The percentage of votes for the Republican in 2020")) %>% 
  rename(`Variables` = var,
         `Source` = source,
         `Descriptions` = descriptions) %>% 
  kable(booktabs = TRUE)
```

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(n = n(),
            case = mean(cases),
            inf = mean(infrate),
            deaths = mean(deaths),
            mort = mean(mortrate),
            dpc = mean(dpc),
            income = mean(mean_household_income)) %>% 
  rename(`Winning Party` = winning_party,
         `Count` = n,
         `Average Cases` = case,
         `Infection Rate` = inf,
         `Average Deaths` = deaths,
         `Mortality Rate` = mort,
         `DPC` = dpc,
         `Income` = income) %>% 
  kable(booktabs = TRUE)
```

@tbl-covid compares the COVID-19 impacts and income levels across the counties won by each party. Even though Republicans won in approximately five-sixths of the counties, counties that supported Biden reported nearly ten times the average number of COVID-19 cases and deaths compared to those that supported Trump. Despite similar infection and mortality rates between the two groups, there is still a notable higher death-to-case rate in counties where Democrats won. These patterns suggest that the counties voting for Democrats are more severely impacted by COVID-19 and are more population-intensive than those voting for the Republicans. Even though only around 500 counties support Democrats, they are densely populated urban areas and wealthier, so Biden can have more electoral votes. We can verify this from income levels, where the counties voting Democrat has a higher mean income.

@@

p1 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = pct_vote_rep), color = "black") +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "grey") +
  labs(title = "The relative share of votes between Republican and Democrat party",
       fill = 'Republican share') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.9)

@@

We can validate our guess from @fig-map, which compares the relative ratio of votes for the two parties and the death to cases rate in maps. From the maps, it is clear more counties had a preference for the Republicans, but the counties with higher death-per-case rates show a preference for Democrats. These counties, such as California and New York, are mostly more affluent. In contrast, the states that are less impacted by COVID-19, such as Utah, vote more for Republicans. The maps match our findings from @tbl-covid.

@@

Given the insights from @tbl-covid and @fig-map, @fig-dpc-income delves deeper into the impact of COVID and income levels on the voting behaviours of the two parties. The left side of @fig-dpc-income displays the share of votes, while the right side focuses on the total number of votes, with each party represented by distinct colours. In terms of the percentage of votes, in general, there is a positive relationship between the DPC and votes for the Democrats but a negative one for the Republicans. However, the counties with a medium value of DPC seem less sensitive to the voting for the two parties, whereas the support of counties at the 'tails' shifts more significantly with changes in the DPC. Meanwhile, there are clear quadratic patterns between the vote for the two parties and income. The lower-income counties tend to favour Republicans, but as income increases, a greater number of counties lean towards voting for Biden rather than Trump.

In addition, despite the average share of votes for Republicans being consistently above 0.5, it does not imply a universal loss for Democrats across all counties. As detailed in @tbl-covid, we know that the number of counties won by Republicans is approximately five times greater than those won by Democrats. With Republicans dominating more counties, they naturally have a higher average share of votes. This pattern also indicates that Democrats usually won the densely populated, urban areas, but Republicans won in more sparsely populated and rural regions [@Spencer2023ElectionFactCheck]. The high population density areas have more votes, which explains why Biden beat Trump. We can verify this from the right side where the number of votes for Biden exceeds Trump as DPC and income level increases. This pattern is particularly pronounced in the wealthiest top 50% of counties and those with DPCs exceeding the 75th percentile.

```{r}
#| fig-cap: The correlation between DPC rate with proportons of people with private insurance and without health insurance for the two parties
#| label: fig-insurance

plot1 <- ggplot(data, aes(x = private_insurance)) +
  #geom_point(aes(y = dpc, colour = winning_party), alpha = 0.25) +  
  geom_smooth(aes(y = dpc, colour = winning_party), 
              method = "lm", formula = y ~ poly(x, 1), se = FALSE) +
  #ylim(1000, 4000) +
  theme_minimal() +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Winning Party:",  
                      labels = c("Democrat", "Republican")) +
  theme(legend.position = "none") + 
  labs(x = "Private Insurance(%)", 
       y = "DPC rate")

plot2 <- ggplot(data, aes(x = no_insurance)) +
  #geom_point(aes(y = dpc, colour = winning_party), alpha = 0.25) +  
  geom_smooth(aes(y = dpc, colour = winning_party), 
              method = "lm", formula = y ~ poly(x, 1), se = FALSE) +
  #ylim(1000, 4000) +
  theme_minimal() +
  scale_colour_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                      name = "Party",  
                      labels = c("Democrat", "Republican")) +
  theme(legend.position = "none") + 
  labs(x = "No Health Insurance(%)", 
       y = "DPC rate")

legend <- get_legend(plot1 + theme(legend.position = "bottom"))
combined_plot <- plot_grid(plot1, plot2, ncol = 2)
combined_plot_with_legend <- plot_grid(combined_plot, legend, ncol = 1, rel_heights = c(1, 0.1))

@fig-insurance shows the correlation between the DPC rate and the proportion of people with private and without health insurance for the two parties. Unsurprisingly, the graph reveals an inverse correlation between the DPC rate and private health insurance – as private insurance coverage increases, the DPC rate tends to decrease. Notably, counties that favoured the Democratic Party exhibit a higher proportion of individuals with private insurance compared to Republican-leaning counties. In contrast, in Republican-dominated counties, there is a pronounced positive correlation between the DPC rate and the lack of health insurance. However, it is relatively flat for Democratic counties. These findings suggest that counties voting Democratic generally have more comprehensive insurance coverage, which appears to mitigate the impact of DPC rates to a greater extent than in Republican counties.

```{r}
#| fig-cap: The distribution of share of White population and high-education across different levels
#| label: fig-bar
data_we <- data %>%
  mutate(white_pct_group = ntile(white_pct, 10),
         prop_education_group = ntile(prop_higher_education, 10))

data_long <- data_we %>%
  pivot_longer(cols = c(pct_vote_rep, pct_vote_demo), names_to = "party", values_to = "votes")

# Replace party names for plotting
data_long$party <- recode(data_long$party, pct_vote_rep = "Republican", pct_vote_demo = "Democrat")

# Plot for white_pct_group
plot1 <- ggplot(data_long, aes(x = factor(white_pct_group), y = votes, group = party)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), aes(fill = party)) +
  scale_fill_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF"),
                    name = "Party") +
  labs(x = "Ratio of White Pop. (Deciles)", y = "Mean Votes(%)", fill = "Party") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot for prop_education_group
plot2 <- ggplot(data_long, aes(x = factor(prop_education_group), y = votes, group = party)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), aes(fill = party)) +
  scale_fill_manual(values = c("Republican" = "#FF9999", "Democrat" = "#9999FF")) +
  labs(x = "Ratio of High-Edu. (Deciles)", y = "Mean Votes(%)", fill = "Party") +
  theme_minimal() +
  theme(legend.position = "none")

legend <- get_legend(plot1 + theme(legend.position = "bottom"))
combined_plot <- plot_grid(plot1, plot2, ncol = 2)
combined_plot_with_legend <- plot_grid(combined_plot, legend, ncol = 1, rel_heights = c(1, 0.1))

@fig-bar compares the ratio of the White population (race diversity) and education levels for the two parties. Interestingly, it suggests that counties with higher racial diversity are less likely to vote for Biden, as opposed to the trend observed for Republican-leaning counties, where increased diversity aligns with increased support. This pattern implies that counties that predominantly vote for Democrats have a higher proportion of White residents. This may explain why they have relatively higher income levels. In terms of the ratio of people with high education attainment, since the average vote for Republicans is always higher than for Democrats, this may indicate that more educated people are likely to vote for Trump.

# Methods{#sec-meethods}

This paper aims to conduct the causal inference between COVID-19 and Trump's loss in 2020. The primary method implemented in this study will be the Propensity Score Matching (PSM). After finding the causal inference, this paper will also conduct a counterfactual analysis to see whether Trump can re-elect without COVID-19.

@@

The treatment refers to the intervention or exposure being studied to understand its potential impact on the outcome [@WikiPropensity2023a]. In other words, we implement specific interventions for a group of people but not for the rest. The group of people receiving the treatment is called the treatment group, and the control group for the rest. For instance, Austin [@Austin2011PropensityScore] examined the effectiveness of a new medical treatment called clampless off-pump coronary artery bypass (clampless OPCAB) to see if it can reduce the in-hospital mortality rate compared to the traditional method. Here, 'treatment' refers to the new surgical procedure, and patients who took this new surgery are the treatment group. They compared the mortality rates between the two groups and concluded that the new treatment could statistically lower the stroke and mortality rate.
